---
title: "Tiny Project: Workforce Analytics and Compensation Drivers"
author: "Fairy"
date: "2026-09-15"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

In human resource analytics, compensation structure is closely tied to professional experience and department roles. This **Tiny Project** investigates the distribution of compensation and organizational demographics across a sample corporate cohort. The analysis includes data importing, structural validation, descriptive statistical measures, categorical visualizations, and linear correlation modeling.

---

# Database / Data Importing

The dataset is imported from the project directory using a relative file path for cross-environment reproducibility.

```{r import-data}
# Read corporate payroll records
employee_data <- read.csv("employee_data.csv")

# Print complete data frame
employee_data

# Inspect preview observations
head(employee_data)

# Structure and attribute datatype check
str(employee_data)
```

**R Code Explanation:**
* `read.csv("employee_data.csv")`: Ingests the tabular comma-separated values into an R data frame.
* `head(employee_data)`: Inspects the top 6 entries to confirm column and row alignments.
* `str(employee_data)`: Displays internal architecture, confirming text categories (`Name`, `Department`, `Performance`) and integer vectors (`Experience`, `Salary`).

---

# Data Preprocessing and Cleaning

The dataset is reviewed to ensure zero missing entries and to verify value ranges before statistical analysis.

```{r data-cleaning}
# Verify presence of missing values (NA)
colSums(is.na(employee_data))

# Generate statistical distribution summary
summary(employee_data)
```

**Interpretation of Preprocessing Results:**
* `colSums(is.na(employee_data))` returns 0 across all variables, confirming complete data integrity with no missing records.
* Experience ranges from 1 year to 9 years (mean: 5.0 years). Monthly salary spans from ₹32,000 to ₹92,000, averaging ₹63,400.

---

# Data Analysis and Statistical Exploration

Central tendency and dispersion are calculated for salaries, followed by frequency distribution tables for categorical factors.

```{r statistical-exploration}
# Statistical measures for Salary
salary_median <- median(employee_data$Salary)
salary_var    <- var(employee_data$Salary)
salary_sd     <- sd(employee_data$Salary)

salary_median
salary_var
salary_sd

# Categorical frequency tables
dept_count   <- table(employee_data$Department)
perf_count   <- table(employee_data$Performance)

dept_count
perf_count
```

**R Code Explanation:**
* `median()` determines the midpoint salary (50th percentile).
* `var()` and `sd()` measure salary dispersion across the cohort.
* `table()` computes frequency breakdowns for qualitative attributes.

**Statistical Interpretation:**
* **Central Tendency:** The median compensation is **₹66,000**, closely matching the group mean.
* **Dispersion:** The standard deviation is **₹20,774.98**, reflecting natural pay disparities tied to seniority.
* **Demographics:** IT and Finance hold the largest shares (3 employees each), while Marketing and HR comprise 2 employees each. Half the team (5 employees) maintains a 'High' performance tier.

---

# Graph and Plot Creation

## Categorical Distribution Visualizations

Pie charts display corporate department representation and employee performance tiers.

```{r pie-charts, fig.width=9, fig.height=4.5}
par(mfrow = c(1, 2))

# Department Representation
pie(
  dept_count,
  main = "Department Distribution",
  labels = paste(names(dept_count), "-", dept_count),
  col = c("#7fc97f", "#beaed4", "#fdc086", "#ffff99")
)

# Performance Tier Distribution
pie(
  perf_count,
  main = "Performance Tier Breakdown",
  labels = paste(names(perf_count), "-", perf_count),
  col = c("#66c2a5", "#fc8d62", "#8da0cb")
)

par(mfrow = c(1, 1))
```

**Interpretation:**
* The workforce shows balanced operational distribution across core domains.
* 50% of the sample cohort qualifies for high-performance ratings, with 30% medium and 20% developing.

---

## Individual-Level Metric Comparisons

Bar charts depict individual salary allocations and professional experience tenure.

```{r bar-plots, fig.width=9, fig.height=5}
# Salary Bar Plot
barplot(
  employee_data$Salary,
  names.arg = employee_data$Name,
  main = "Monthly Salary by Employee",
  xlab = "Employee",
  ylab = "Salary (INR)",
  col = "#41b6c4",
  las = 2
)

# Experience Bar Plot
barplot(
  employee_data$Experience,
  names.arg = employee_data$Name,
  main = "Years of Experience by Employee",
  xlab = "Employee",
  ylab = "Years",
  col = "#fe9929",
  las = 2
)
```

**Interpretation:**
* Ramesh (₹92,000; 9 years) and Manish (₹88,000; 8 years) command the highest compensation packages.
* Junior team members Ananya (₹32,000; 1 year) and Kavita (₹38,000; 2 years) occupy baseline compensation levels.

---

## Bivariate Correlation Analysis

Evaluating the association between professional experience (tenure) and salary compensation using Pearson's correlation and ordinary least squares linear regression.

```{r correlation-scatter, fig.width=7, fig.height=5}
# Compute Pearson correlation coefficient
correlation_val <- cor(employee_data$Experience, employee_data$Salary)
correlation_val

# Scatter plot of Experience vs. Salary
plot(
  employee_data$Experience,
  employee_data$Salary,
  main = paste("Experience vs Salary (r =", round(correlation_val, 4), ")"),
  xlab = "Experience (Years)",
  ylab = "Salary (INR)",
  pch = 19,
  col = "#225ea8"
)

# Overlay trendline
abline(lm(Salary ~ Experience, data = employee_data), col = "red", lwd = 2)
```

**Code Explanation & Interpretation:**
* `cor()` determines linear correlation strength ($r \approx 0.994$).
* `plot()` and `abline(lm(...))` display observations with an overlaid regression trendline.
* **Interpretation:** A correlation coefficient of **+0.994** demonstrates a near-perfect positive correlation, validating that years of organizational experience serve as the primary predictive driver for base compensation.

---

# Findings and Conclusion

* **Data Integrity:** All observations were clean, non-null, and structurally consistent.
* **Compensation Drivers:** Employee remuneration scales consistently with professional seniority across all departments.
* **Performance Clustering:** Senior professionals with higher pay rates consistently demonstrated higher evaluation ratings.
* **Summary Insight:** Organizational retention and structured experience tiers directly correlate with compensation ladders, showing predictable annual growth per additional year of tenure.